### 2018-04-26 1_intro.tex
Change the name of \mousquetaires.

The first tool that performs the WMM-aware portability analysis is \Porthos introduced in April~2017~\cite{Porthos17}. This tool reduces described problem to a bounded reachability problem, which can be solved with help of an SMT-solver. This approach allows to capture symbolically the semantics of analysing program and both weak memory models into single SMT-formula, augmented by the reachability assertion. As most modern SMT-solvers are efficient enough to be able to operate the state space of size millions of variables bounded by millions of constraints~(\cite{malik2009boolean}), the used method can be applicable in solving the real-world problems.
% TODO: search space size: https://courses.cs.washington.edu/courses/csep573/11wi/lectures/ashish-satsolvers.pdf slide

@@

Current work aims to rework the proof-of-concept tool \Porthos by extending the input language, which currently represents the minimum subset of C, and revising the general architecture of the tool in order to enhance performance, reliability and maintainability.
%interprocedural analysis
We called the new tool the~\tool{mousquetaires}~framework as now it not only performs the portability analysis, but can serve as the basis for SMT-solver driven weak memory model analysis.

>>>>

The first tool that performs the WMM-aware portability analysis is \porthos introduced in April~2017~\cite{Porthos17}. This tool reduces described problem to a bounded reachability problem, which can be solved with help of an SMT-solver. This approach allows to capture symbolically the semantics of analysing program and both weak memory models into single SMT-formula, augmented by the reachability assertion. As most modern SMT-solvers are efficient enough to be able to operate the state space of size millions of variables bounded by millions of constraints~(\cite{malik2009boolean}), the used method can be applicable in solving the real-world problems.
% TODO: search space size: https://courses.cs.washington.edu/courses/csep573/11wi/lectures/ashish-satsolvers.pdf slide

@@

Current work aims to rework the proof-of-concept tool \porthos by extending the input language, which currently represents the minimum subset of C, and revising the general architecture of the tool in order to enhance performance, reliability and maintainability.
%interprocedural analysis
As the general architecture and almost all components of \porthos have been redesigned, the tool received a new name -- \porthos[2].
Considering the enhancements of the architecture, \porthos[2] represents a generalised framework for SMT-based memory model-aware analysis, which can not only perform the portability analysis, but can serve as a basis for other kinds of static code analysis.

### 2018-04-26 2_wmm.tex
Change the name of \mousquetaires.

	\item Events from the same process are related by the \textit{scope relation} \sr~$\subset~\mathbb{E}~\times~\mathbb{E}$.
    In contrast to the \tool{herd} tool, the \mousquetaires does not use hierarchy of scopes (depicted as the scope tree); instead, it uses simple lables that indicate which process has produced certain event.
\end{itemize}

>>>>

	\item Events from the same process are related by the \textit{scope relation} \sr~$\subset~\mathbb{E}~\times~\mathbb{E}$.
    In contrast to the \tool{herd} tool, the \porthos[2] does not use hierarchy of scopes (depicted as the scope tree); instead, it uses simple lables that indicate which process has produced certain event.
\end{itemize}

### 2018-04-26 3_encod.tex
Change the name of \mousquetaires.

While working on the \tool{mousquetaires}, we applied some modifications of the encoding scheme for the control-flow. The changes are conditioned by the need to be able to process an arbitrary control-flow produced by conditional and unconditional jumps of C language. 
For that, we compile the recursive abstract syntax tree~(AST) of the parsed C-code to the plain (non-recursive) event-flow graph.
We show %TODO! to show, actually
that the new encoding is smaller than the old one used in \Porthos since it does not produces new variables for each high-level statement of the input language.
%The \tool{PORTHOS} tool encodes the \textit{instructions}, which have recursive nature, in a recursive manner. This means, it adds synthetic composite instructions for both linear (sequential) and non-linear (branching) instructions. 
For instance, \Porthos uses the encoding scheme where the control-flow of the sequential instruction $i_1 = i_2; i_3$ was encoded as
$\phi_{CF}(i_2;i_3) = (cf_{i_1} \Leftrightarrow (cf_{i_2} \land cf_{i_3})) \land \phi_{CF}(i_2) \land \phi_{CF}(i_3)$,
and control-flow of the branching instruction $i_1 = (c \, ? \, i_2 \, : \, i_3)$ was encoded as
$\phi_{CF}( \, c \, ? \, i_2 \, : \, i_3) = (cf_{i_1} \Leftrightarrow (cf_{i_2} \lor cf_{i_3})) \land \phi_{CF}(i_2) \land \phi_{CF}(i_3)$
(here we used the notation of C-like ternary operator \texttt{x?y:z} for defining the conditional expression $\mathtt{if} x \mathtt{then} y \mathtt{else} z$).
In contrast, the new scheme implemented in \tool{mousquetaires} firstly compiles the recursive high-level code into the linear low-level event-based representation, that is then encoded into an SMT-formula. The encoding of branching nodes depends on the \textit{guards}, the value of conditional variable on the branching state, which in turn is encoded as data-flow constraint (see Section~\ref{ch:port:enc:df}).

@@

To encode the data-flow constraints, we use the \textit{static single-assignment (SSA) form} in order to be able to capture an arbitrary data-flow into a single SMT-formula.
The SSA form requires each variable to be assigned only once within entire program.
In contrast, \Porthos used the dynamic single-assignment (DSA) form, that requires indices to be unique within a branch.
Although the number of variable references (each of which is encoded as unique SMT-variable) on average is logarithmically less in case of the DSA form than the SSA form, the result SMT-formula still needs to be complemented by same number of equality assertions when encoding the data-flow in merge points~\cite{Porthos17}.

@@

The time of described algorithm is linear of the size of event-flow graph since it performs only single traverse of the graph.
%Notwithstanding the overhead of storing (or, equivalently, computing with %TODO: are you sure that 'with'?
%linear time) the predecessors of each node in order to be able to convert the program into an SSA form, the time of such a transformation reduced %TODO: How much ????
%comparing to the algorithm implemented in \Porthos, which recomputed SSA-indices recursively for each instruction. %TODO: PROOF OR REMOVE !  Maybe, some words of copying, merging SSA maps ?

As it has been described before, the \rf-relation links data-flow between events %TODO: another word?
of data-flow stored in equivalence assertions over the SSA-variables. 
%The encoding of the \rf-relation, which links the SSA-variables to the original variables, 
The encoding of this linkage left untouched as it is implemented in \Porthos: for each pair of events $e_1$ and $e_2$ linked by the \rf-relation, we add the following constraint:

>>>>

While working on the \tool{mousquetaires}, we applied some modifications of the encoding scheme for the control-flow. The changes are conditioned by the need to be able to process an arbitrary control-flow produced by conditional and unconditional jumps of C language. 
For that, we compile the recursive abstract syntax tree~(AST) of the parsed C-code to the plain (non-recursive) event-flow graph.
We show %TODO! to show, actually
that the new encoding is smaller than the old one used in \porthos since it does not produces new variables for each high-level statement of the input language.
%The \tool{PORTHOS} tool encodes the \textit{instructions}, which have recursive nature, in a recursive manner. This means, it adds synthetic composite instructions for both linear (sequential) and non-linear (branching) instructions. 
For instance, \porthos uses the encoding scheme where the control-flow of the sequential instruction $i_1 = i_2; i_3$ was encoded as
$\phi_{CF}(i_2;i_3) = (cf_{i_1} \Leftrightarrow (cf_{i_2} \land cf_{i_3})) \land \phi_{CF}(i_2) \land \phi_{CF}(i_3)$,
and control-flow of the branching instruction $i_1 = (c \, ? \, i_2 \, : \, i_3)$ was encoded as
$\phi_{CF}( \, c \, ? \, i_2 \, : \, i_3) = (cf_{i_1} \Leftrightarrow (cf_{i_2} \lor cf_{i_3})) \land \phi_{CF}(i_2) \land \phi_{CF}(i_3)$
(here we used the notation of C-like ternary operator \texttt{x?y:z} for defining the conditional expression $\mathtt{if} x \mathtt{then} y \mathtt{else} z$).
In contrast, the new scheme implemented in \tool{mousquetaires} firstly compiles the recursive high-level code into the linear low-level event-based representation, that is then encoded into an SMT-formula. The encoding of branching nodes depends on the \textit{guards}, the value of conditional variable on the branching state, which in turn is encoded as data-flow constraint (see Section~\ref{ch:port:enc:df}).

@@

To encode the data-flow constraints, we use the \textit{static single-assignment (SSA) form} in order to be able to capture an arbitrary data-flow into a single SMT-formula.
The SSA form requires each variable to be assigned only once within entire program.
In contrast, \porthos used the dynamic single-assignment (DSA) form, that requires indices to be unique within a branch.
Although the number of variable references (each of which is encoded as unique SMT-variable) on average is logarithmically less in case of the DSA form than the SSA form, the result SMT-formula still needs to be complemented by same number of equality assertions when encoding the data-flow in merge points~\cite{Porthos17}.

@@

The time of described algorithm is linear of the size of event-flow graph since it performs only single traverse of the graph.
%Notwithstanding the overhead of storing (or, equivalently, computing with %TODO: are you sure that 'with'?
%linear time) the predecessors of each node in order to be able to convert the program into an SSA form, the time of such a transformation reduced %TODO: How much ????
%comparing to the algorithm implemented in \porthos, which recomputed SSA-indices recursively for each instruction. %TODO: PROOF OR REMOVE !  Maybe, some words of copying, merging SSA maps ?

As it has been described before, the \rf-relation links data-flow between events %TODO: another word?
of data-flow stored in equivalence assertions over the SSA-variables. 
%The encoding of the \rf-relation, which links the SSA-variables to the original variables, 
The encoding of this linkage left untouched as it is implemented in \porthos: for each pair of events $e_1$ and $e_2$ linked by the \rf-relation, we add the following constraint:

### 2018-04-26 4_input.tex
Change the name of \mousquetaires.

The \Porthos tool used a small subset of C as an input language as described on Figure~\ref{syntax:porthos}\cite{Porthos17}:

\begin{figure}[H]
\begin{multicols}{2}
\begin{lstlisting}[mathescape=true,
        caption={Syntax of an input language of \Porthos},
        % captionpos=b,
        label={syntax:porthos},
        % basicstyle=\ttfamily\scriptsize,
        % keywordstyle=\bfseries\color{blue},
        %literate={<--}{$\leftarrow{}${}1},%{::=}{???},%:\!:\!=},
        %morekeywords={if,then,else,return,while,program,thread,mfence,sync,lwsync,isync}
        %alsoletter={:,|,;},%
        %morekeywords={:,|,;},%
        %keywordstyle={\color{red}},% 
        breaklines=true,
        % xleftmargin=5mm,framexleftmargin=5mm
        ]
<prog>  : program <thrd>*
<thrd>  : thread <tid> <inst>
<inst>  : <atom> 
    | <inst> ; <inst>
    | while <pred> <inst>
    | if <pred > then <inst> else <inst>
<atom>  : <reg> <-- <exp> 
    | <reg> <-- <loc>
    | <loc> := <reg> 
    | mfence
    | sync
    | lwsync
    | isync
\end{lstlisting}
%\caption={Syntax of an input language of \Porthos}
%\label={syntax:porthos}
\end{multicols}
\end{figure}

// todo: old syntax more precisely

The input language of \Porthos supported recursive definitions of control-flow instructions, which needed to be extended to support arbitrary jumps expressible in C. Also, for the purpose of simplicity, all kinds of fence instructions and different types of variable assignments (shared variable assignment `\texttt{:=}' or local variable assignment `$\mathtt{\rightarrow}$') were defined directly in the grammar for \Porthos, whereas the semantics of function invocations need to be determined dynamically during the interpretation
% according to the set of known functions
(see Chapter~\ref{???}).

//todo: new syntax

>>>>

%TODO: say sth about 100500 litmus-tests for kernel

The previous version of \porthos processed the small subset of C as an input language (see grammar in Figure~\ref{syntax:in_grammar_pts})~\cite{Porthos17}.
It supported recursive definitions of control-flow instructions, that were directly encoded into SMT-formula.
The grammar supported by \porthos[2] was extended in order to support expressions valid in C. %TODO: examples, how imenno extended
Also, the input language for the first version of \porthos  
Also, for the purpose of simplicity, all kinds of fence instructions and different types of variable assignments (shared variable assignment~`\texttt{:=}' or local variable assignment~`$\mathtt{\leftarrow}$') were defined directly in the grammar for \porthos, whereas the semantics of function invocations need to be determined dynamically during the interpretation
% according to the set of known functions
(see Chapter~\ref{???}).

\begin{figure}%[H]
%\begin{multicols}{2}
\begin{lstlisting}[mathescape=true,
        caption={Syntax of an input language of \porthos},
        label={syntax:in_grammar_pts},
        literate={<--}{{$\leftarrow$}}1,
        morekeywords={if,then,else,return,while,program,thread,mfence,sync,lwsync,isync}
        breaklines=true,
        basicstyle=\ttfamily\scriptsize,
        ]
<prog> : program <thrd>*
       ;
<thrd> : thread <tid> <inst>
       ;
<inst> : <atom> 
       | <inst> ; <inst>
       | while <pred> <inst>
       | if <pred > then <inst> else <inst>
       ;
<atom> : <reg> <-- <exp> 
       | <reg> <-- <loc>
       | <loc> := <reg> 
       | mfence
       | sync
       | lwsync
       | isync
       ;
\end{lstlisting}
%\end{multicols}
\end{figure}

Both \porthos and \porthos[2] \ use the parser generator ANTLR~\cite{parr2013definitive}, the powerfull language processing tool. The (hand-written) ANTLR grammar used in \porthos is available at Appendix~\ref{apx:in_grammar_pts}, whereas the \porthos[2] uses the grammar of C language proposed in the C11 standard~\cite{jtc2011sc22} (the ANTLR grammar can be found in the official ANTLR repository on GitHub~\footnote{\url{https://github.com/antlr/grammars-v4}}).

### 2018-04-26 5_impl.tex
Change the name of \mousquetaires.

\chapter{\nohyphens{The~\mousquetaires:~implementation}}
\label{ch:impl}

@@

This Chapter describes the architecture of the \mousquetaires framework.

The programming language choice for \mousquetaires was made in favour of java in order to reuse some parts of its predecessor \Porthos that was written in java. However this language does not show best results in performance benchmarks (comparing to, for example, C++)~\cite{TODO}, the performance cornerstone of \mousquetaires (as well as any other SMT-based code analyser) is the phase of solving the SMT-formula, which is left to the third-party SMT-solver called from \mousquetaires via java API. However, considering the perspective of using \mousquetaires as a static analyser for real-world programs, we also must to take into account memory problems at both encoding and solving stages.

@@

The main reason for starting the work on re-implementing the \Porthos tool was the need to extend input language to be able to process real C programs in perspective. The existing \Porthos architecture mixes high-level recursive \textit{instructions} (statements of C) with low-level non-recursive \textit{events}, to which the high-level instructions are compiled (see classes of package `\texttt{dartagnan.program}' in the \Porthos project~\footnote{Project web site: \url{http://github.com/hernanponcedeleon/PORTHOS}}). Both instructions and events are then encoded into the SMT-formula. However, this approach is poorly extensible as to adding support for a high-level control-flow instruction (say, the \texttt{do-while} loop in C) requires making changes in many places of code from parser to encoder. Moreover, we encounter serious problems when we need to add support for control-flow jumps (as \texttt{continue}, \texttt{break}, \texttt{goto} in C).

Therefore, while planning the architecture of \mousquetaires, we decided to clearly separate high-level recursive code representation as Abstract Syntax Tree (AST) and low-level compiled event-based representation as a flow-graph. Such a modular architecture allows to use multiple input-language parsers and convert parsed syntax trees to our AST, thus having support for multiple languages (for instance, the original input language of \Porthos tool, two syntaxes of litmus tests used by \tool{herd}, and even assembly language for any supported architecture).

@@

Generally, \mousquetaires uses following  ...

>>>>

\chapter{\nohyphens{The~\porthos[2]:~implementation}}
\label{ch:impl}

@@

This Chapter describes the architecture of the \porthos[2] framework.

The programming language choice for \porthos[2] was made in favour of java in order to reuse some parts of its predecessor \porthos that was written in java. However this language does not show best results in performance benchmarks (comparing to, for example, C++)~\cite{TODO}, the performance cornerstone of \porthos[2] (as well as any other SMT-based code analyser) is the phase of solving the SMT-formula, which is left to the third-party SMT-solver called from \porthos[2] via java API. However, considering the perspective of using \porthos[2] as a static analyser for real-world programs, we also must to take into account memory problems at both encoding and solving stages.

@@

The main reason for starting the work on re-implementing the \porthos tool was the need to extend input language to be able to process real C programs in perspective. The existing \porthos architecture mixes high-level recursive \textit{instructions} (statements of C) with low-level non-recursive \textit{events}, to which the high-level instructions are compiled (see classes of package `\texttt{dartagnan.program}' in the \porthos project~\footnote{Project web site: \url{http://github.com/hernanponcedeleon/PORTHOS}}). Both instructions and events are then encoded into the SMT-formula. However, this approach is poorly extensible as to adding support for a high-level control-flow instruction (say, the \texttt{do-while} loop in C) requires making changes in many places of code from parser to encoder. Moreover, we encounter serious problems when we need to add support for control-flow jumps (as \texttt{continue}, \texttt{break}, \texttt{goto} in C).

Therefore, while planning the architecture of \porthos[2], we decided to clearly separate high-level recursive code representation as Abstract Syntax Tree (AST) and low-level compiled event-based representation as a flow-graph. Such a modular architecture allows to use multiple input-language parsers and convert parsed syntax trees to our AST, thus having support for multiple languages (for instance, the original input language of \porthos tool, two syntaxes of litmus tests used by \tool{herd}, and even assembly language for any supported architecture).

@@

Generally, \porthos[2] uses following  ...
